### 2014-05-20 ApTranscriptome_TR.Rmd
Updates to TR

**May 16, 2014**

@@

# from either AWS or GoogleDrive
annotationURL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
#a2 <- getURL("https://googledrive.com/host/0B75IymziRJ_9Tlg1U1Vxbjk1bzg") # GoogleDrive link

annotationfile <- read.csv(textConnection(annotationURL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotationfile)

# Convert to data.table
annotationtable <- data.table(annotationfile)
head(annotationtable)
```

@@

```{r extractionTranscripts}
A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
Ar.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]

A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
Ar.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]

A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
Ar.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]

A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
Ar.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
```

@@

t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
```

Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.

@@

```{r T_on_low}
# transcripts expressed at *Low* temperatures in A22
A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.low, "A22")]
str(A22.low.TPM.dt.sub)

@@

for(i in unique(A22.low.TPM.dt.sub$Transcript)) {
    subdf <- A22.low.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val < 21), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    A22.low.T_on[which(A22.low.T_on$Transcript == i), "T_on"] <- T_on
}

# repeat for Ar
Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.low, "Ar")]
str(Ar.low.TPM.dt.sub)

@@

t.test(Ar.low.T_on$T_on, A22.low.T_on$T_on)
```

@@

```{r gene_overlap}
# transcripts that are low in A22 and intermediate in Ar
A22.low.Ar.int <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
A22.low.Ar.int.lms <- responsive.lms[A22.low.Ar.int]

# transcripts that are low in both A22 and Ar
A22.low.Ar.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
A22.low.Ar.low.lms <- responsive.lms[A22.low.Ar.low]

head(int.low.overlap)

# compare linear and quadratic model terms

# transcripts where TPM increases at low temps & decreases at high temps
A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low,Ar.int), "A22")]

for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
    s <- A22.low.Ar.int.TPM.dt.sub[i]
    X11()
    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
    points(s$val, s$TPM, pch = 8, col = "red")
    if(min(s[which(s$val < 30) , pTPM]) < s[which(s$val == 38.5), pTPM]) print("On low") else {
        print("Off high")
    }
}

# A22 low Ar int
A22.low.Ar.int.low.type <- ddply(A22.low.Ar.int.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))

# transcripts where TPM increases at low temps & decreases at high temps

A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low, Ar.int), "A22")]
dim(A22.low.Ar.int.TPM.dt.sub)

l5 <- length(unique(A22.low.Ar.int.TPM.dt.sub$Transcript))

A22.low.Ar.int.T_on <- data.frame(Transcript = unique(A22.low.Ar.int.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l5), type = rep("High", length = l5), T_on = NA)

# of these, how many have greatest change in expression as a decrease at high temperatures?
for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)) {
    subdf <- A22.low.Ar.int.TPM.dt.sub[i]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    A22.low.Ar.int.T_on[which(A22.low.Ar.int.T_on$Transcript == i), "T_on"] <- T_on
}

```

@@

```{r annotation}
setkey(annotationtable, Sequence.Name)
signif.transcripts <- data.table(signif.transcripts)
setkey(signif.transcripts, Transcript)
signif.transcripts <- annotationtable[signif.transcripts]
setnames(signif.transcripts, "Sequence.Name", "Transcript")
```

@@

# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.lms compared to full list
resp.stress.responsive.count <- nrow(responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.lms.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)

@@

hsp_all <- annotationtable[grep("shock", annotationtable$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
nrow(hsp_all)
```

@@

```{r geneid2go, echo=TRUE, eval=TRUE, cache=TRUE}
# create geneid2go.map file from FastAnnotator AnnotationTable.txt
geneid2GOmap(annotationfile)
```

@@

```{r gsea_setdiff, results = 'hide'}
# A22 'High' genes not in Ar**
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)

# *Ar 'High' genes not in A22**
Ar.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.high) <- names(Ap.geneList)
Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high))] <- 1
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)

# A22 'Low' genes not in Ar**
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low))] <- 1
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
# Ar 'Low' genes not in A22**
Ar.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.low) <- names(Ap.geneList)
Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low))] <- 1
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)

# A22 'Bimodal' genes not in Ar**
A22.geneList.bim <- rep(0, length = length(Ap.geneList))
names(A22.geneList.bim) <- names(Ap.geneList)
A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim))] <- 1
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)

# Ar 'Bimodal' genes not in A22**
Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.bim) <- names(Ap.geneList)
Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim))] <- 1
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)

# A22 'Intermediate' genes not in Ar**
A22.geneList.int <- rep(0, length = length(Ap.geneList))
names(A22.geneList.int) <- names(Ap.geneList)
A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int))] <- 1
table(A22.geneList.int)
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)

# Ar 'Intermediate' genes not in A22**
Ar.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.int) <- names(Ap.geneList)
Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int))] <- 1
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)

@@

```{r ggplot_high}
trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony2) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp_A22_high)

```

>>>>

**May 20, 2014**

@@

# from either AWS or GoogleDrive
annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
#a2 <- getURL("https://googledrive.com/host/0B75IymziRJ_9Tlg1U1Vxbjk1bzg") # GoogleDrive link

annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotation.file)

# Convert to data.table
annotation.table <- data.table(annotation.file)
head(annotation.table)
```

@@

```{r extract_Transcripts}
A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
Ar.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]

A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
Ar.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]

A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
Ar.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]

A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
Ar.int <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
```

@@

(T_on.high.ttest <- t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on))
```

Genes with increased expression at *High* temperatures are on average turned on 2.5°C higher in *ApVT* than *AcNC*.

@@

```{r T_on_low}
# transcripts expressed at *Low* temperatures in A22
A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.low, A22.bim), "A22")]
setkey(A22.low.TPM.dt.sub, Transcript)
str(A22.low.TPM.dt.sub)

@@

# loop across transcripts, calculating T_on
for(i in unique(A22.low.TPM.dt.sub$Transcript)) {
    subdf <- A22.low.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val < 21), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    A22.low.T_on[which(A22.low.T_on$Transcript == i), "T_on"] <- T_on
}

# repeat for Ar
Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.low, Ar.bim), "A22")]
setkey(Ar.low.TPM.dt.sub, Transcript)
str(Ar.low.TPM.dt.sub)

@@

(T_on.low.ttest <- t.test(Ar.low.T_on$T_on, A22.low.T_on$T_on))
```

@@

@@

```{r annotation}
setkey(annotation.table, Sequence.Name)
signif.transcripts <- data.table(signif.transcripts)
setkey(signif.transcripts, Transcript)
signif.transcripts <- annotation.table[signif.transcripts]
setnames(signif.transcripts, "Sequence.Name", "Transcript")
```

@@

# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.lms compared to full list
resp.stress.responsive.count <- nrow(responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotation.table[grep("GO:0006950", annotation.table$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.lms.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotation.table) - resp.stress.all.count), nrow = 2)

@@

hsp_all <- annotation.table[grep("shock", annotation.table$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
nrow(hsp_all)
```

@@

```{r geneid2go, echo=TRUE, eval=TRUE, cache=TRUE}
# create geneid2go.map file from FastAnnotator annotationtable.txt
geneid2GOmap(annotation.file)
```

@@

```{r extract_Transcripts}
A22.only.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
Ar.only.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]

A22.only.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
Ar.only.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]

A22.only.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
Ar.only.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]

A22.only.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
Ar.only.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
```

```{r gsea_setdiff, results = 'hide'}
# A22 'High' genes not in Ar**
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.only.high))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)

# *Ar 'High' genes not in A22**
Ar.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.high) <- names(Ap.geneList)
Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.only.high))] <- 1
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)

# A22 'Low' genes not in Ar**
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.only.low))] <- 1
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
# Ar 'Low' genes not in A22**
Ar.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.low) <- names(Ap.geneList)
Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.only.low))] <- 1
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)

# A22 'Bimodal' genes not in Ar**
A22.geneList.bim <- rep(0, length = length(Ap.geneList))
names(A22.geneList.bim) <- names(Ap.geneList)
A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.only.bim))] <- 1
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)

# Ar 'Bimodal' genes not in A22**
Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.bim) <- names(Ap.geneList)
Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.only.bim))] <- 1
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)

# A22 'Intermediate' genes not in Ar**
A22.geneList.int <- rep(0, length = length(Ap.geneList))
names(A22.geneList.int) <- names(Ap.geneList)
A22.geneList.int[(which(names(A22.geneList.int) %in% A22.only.int))] <- 1
table(A22.geneList.int)
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)

# Ar 'Intermediate' genes not in A22**
Ar.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.int) <- names(Ap.geneList)
Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.only.int))] <- 1
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)

@@

```{r ggplot_high}
trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony2) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp_A22_high)
```

An observation from GO analysis is that there is an overlap between enriched terms in the *Low* group for *A22* and the *Intermediate* group for *Ar*. A potential explanation for this pattern is if the *Low* genes in *A22* shut down at high temperatures, but not low temperatures, rather than are up-regulated at low temperatures.

```{r gene_overlap}
# transcripts where TPM increases at low temps & decreases at high temps
A22.low.Ar.int.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(intersect(A22.low,Ar.int), "A22")]

for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
    s <- A22.low.Ar.int.TPM.dt.sub[i]
    X11()
    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
    points(s$val, s$TPM, pch = 8, col = "red")
    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh") else {
        print("OnLow")
    }
}

#    if(min(s[which(s$val < 30) , pTPM]) < s[which(s$val == 38.5), pTPM] & s[which(s$val == 0), pTPM] < (s[which(s$val == 21), pTPM] + sd(s$pTPM))) print("On low") else {

# tabulate the number of transcripts that have lowest expression at maximum temperature "OffHigh"
A22.low.Ar.int.low.type <- ddply(A22.low.Ar.int.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))

A22.low.Ar.low.low.type <- ddply(A22.low.Ar.low.TPM.dt.sub, .(Transcript), summarise, lt = ifelse(pTPM[which(val == 38.5)] == min(pTPM), "OffHigh", "OnLow"))

# table results and perform chi-square test
low.table <- rbind(LowInt = table(A22.low.Ar.int.low.type$lt), LowLow = table(A22.low.Ar.low.low.type$lt))
low.table
chisq.test(low.table)
```
